---
title: "crime_analysis2"
output:
  pdf_document: default
  html_document: default
date: "2024-03-27"
---

# Load in Data

```{r}
library(readxl)
library(ggplot2)
library(dplyr)
library(reshape2)
```

```{r}

# Specify the file path
file_path <- "C:/Users/jans7/OneDrive - Marquette University/SP24/COSC 6510 - Data Intelligence/KC_Schools.xlsx"

# Read in the Excel file
school_data <- read_excel(file_path)

# View the first few rows of the data
(school_data)
```


```{r}

# Specify the file path
file_path <- "C:/Users/jans7/OneDrive - Marquette University/SP24/COSC 6510 - Data Intelligence/crimedata_clean.csv"

crime_data_clean <- read.csv(file_path)

# View the first few rows of the data
head(crime_data_clean)
```


```{r}

# Specify the file path
file_path <- "C:/Users/jans7/OneDrive - Marquette University/SP24/COSC 6510 - Data Intelligence/school_crimes.csv"

school_crimes <- read.csv(file_path)

# View the first few rows of the data
head(school_crimes)
```



# KC Schools Analysis 

## Has Crime increased or decreased near Schools?

```{r}
head(school_crimes)
```

Changing the orientation of the data so it makes it easier for analysis

```{r}
start_date <- as.Date("2015-01-01")
end_date <- as.Date("2023-12-31")
months <- seq(start_date, end_date, by = "month")

year_month_df <- data.frame(Year = format(months, "%Y"), Month = format(months, "%B"))

# Transpose the dataframe
transposed_df <- t(school_crimes[, -c(1:8)])  # Exclude the first column (school names) before transposing
colnames(transposed_df) <- school_crimes[, 2]
transposed_df <- as.data.frame(transposed_df)
final_df <- cbind(year_month_df, transposed_df)

head(final_df)
tail(final_df)

```

fixing format of columns

```{r}
# Convert Year and Month columns to numeric
final_df$Year <- as.numeric(final_df$Year)
month_mapping <- c("January" = 1, "February" = 2, "March" = 3, "April" = 4,
                   "May" = 5, "June" = 6, "July" = 7, "August" = 8,
                   "September" = 9, "October" = 10, "November" = 11, "December" = 12)
final_df$Month <- month_mapping[final_df$Month]
```


```{r}
# to make the plot look good can only do 1 school at a time

school_name <- "Central High School"  # Replace with the desired school name


school_df <- final_df[, c("Year", "Month", school_name)]

# Plot the data for the selected school
ggplot(school_df, aes(x = Month, y = !!rlang::sym(school_name), color = factor(Year))) +
  geom_line() +
  labs(x = "Month", y = "Number of Crimes", title = paste("Crimes per Month for", school_name)) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

now lets plot average as well as a confidence interval just so you can see if there has been a change

```{r}
# to make the plot look good can only do 1 school at a time
school_name <- "Central High School"  # Replace with the desired school name
school_df <- final_df[, c("Year", "Month", school_name)]

# Calculate average and standard deviation across all years
summary_stats <- school_df %>%
  group_by(Month) %>%
  summarise(avg = mean(!!rlang::sym(school_name)), 
            sd = sd(!!rlang::sym(school_name)),
            school = school_name)

# Plot 
ggplot() +
  geom_line(data = school_df, aes(x = Month, y = !!rlang::sym(school_name), color = factor(Year))) +
  geom_line(data = summary_stats, aes(x = Month, y = avg, group = school), color = "black", linetype = "dashed", size = 1) + # Add average line
  geom_ribbon(data = summary_stats, aes(x = Month, ymin = avg - sd, ymax = avg + sd, group = school), fill = "grey70", alpha = 0.3) + # Add confidence interval
  labs(x = "Month", y = "Number of Crimes", title = paste("Crimes per Month for", school_name)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = 1:12) + # Set x-axis tick marks from 1 to 12
  scale_color_discrete(name = "Year") # Customize legend label
```
now looking at two other schools for comparison

```{r}
# to make the plot look good can only do 1 school at a time
school_name <- "Border Star Montessori School"  # Replace with the desired school name
school_df <- final_df[, c("Year", "Month", school_name)]

# Calculate average and standard deviation across all years
summary_stats <- school_df %>%
  group_by(Month) %>%
  summarise(avg = mean(!!rlang::sym(school_name)), 
            sd = sd(!!rlang::sym(school_name)),
            school = school_name)

# Plot 
ggplot() +
  geom_line(data = school_df, aes(x = Month, y = !!rlang::sym(school_name), color = factor(Year))) +
  geom_line(data = summary_stats, aes(x = Month, y = avg, group = school), color = "black", linetype = "dashed", size = 1) + # Add average line
  geom_ribbon(data = summary_stats, aes(x = Month, ymin = avg - sd, ymax = avg + sd, group = school), fill = "grey70", alpha = 0.3) + # Add confidence interval
  labs(x = "Month", y = "Number of Crimes", title = paste("Crimes per Month for", school_name)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = 1:12) + # Set x-axis tick marks from 1 to 12
  scale_color_discrete(name = "Year") # Customize legend label
```

```{r}
# to make the plot look good can only do 1 school at a time
school_name <- "Manual Career and Technical Center"  # Replace with the desired school name
school_df <- final_df[, c("Year", "Month", school_name)]

# Calculate average and standard deviation across all years
summary_stats <- school_df %>%
  group_by(Month) %>%
  summarise(avg = mean(!!rlang::sym(school_name)), 
            sd = sd(!!rlang::sym(school_name)),
            school = school_name)

# Plot 
ggplot() +
  geom_line(data = school_df, aes(x = Month, y = !!rlang::sym(school_name), color = factor(Year))) +
  geom_line(data = summary_stats, aes(x = Month, y = avg, group = school), color = "black", linetype = "dashed", size = 1) + # Add average line
  geom_ribbon(data = summary_stats, aes(x = Month, ymin = avg - sd, ymax = avg + sd, group = school), fill = "grey70", alpha = 0.3) + # Add confidence interval
  labs(x = "Month", y = "Number of Crimes", title = paste("Crimes per Month for", school_name)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = 1:12) + # Set x-axis tick marks from 1 to 12
  scale_color_discrete(name = "Year") # Customize legend label
```

# Total

## Which Schools Have the Most Crime Near them

```{r}

# to store the results
top_schools <- data.frame()


for (year in unique(final_df$Year)) {
  # Filter data for the current year
  year_data <- filter(final_df, Year == year)
  
  # Sum the crimes for each school
  year_sum <- year_data %>%
    summarise(across(3:38, sum)) %>%
    melt(id.vars = NULL, variable.name = "School", value.name = "Total_Crimes") %>%
    arrange(desc(Total_Crimes))
  
  # Select the top 3 schools
  top_3 <- year_sum[1:3, ]
  top_3$Year <- year
  
  top_schools <- bind_rows(top_schools, top_3)
}

print(top_schools)
```

Lets now visualize this

```{r}
ggplot(top_schools, aes(x = as.factor(Year), y = Total_Crimes, fill = School)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Year", y = "Total Number of Crimes", title = "Top 3 Schools with Most Crime Each Year") +
  scale_fill_brewer(palette = "Set3") + 
  theme_minimal() +
  theme(legend.title = element_blank())  
```
This is very useful information 


## Which Schools Have the Least Crime Near them

```{r}

# to store the results
bot_schools <- data.frame()


for (year in unique(final_df$Year)) {
  # Filter data for the current year
  
  year_data <- filter(final_df, final_df$Year == year)
  # Sum the crimes for each school
  year_sum <- year_data %>%
    summarise(across(3:38, sum)) %>%
    melt(id.vars = NULL, variable.name = "School", value.name = "Total_Crimes") %>%
    arrange(Total_Crimes)
  
  # Select the bottom 3 schools
  bot_3 <- year_sum[1:3, ]
  bot_3$Year <- year
  
  bot_schools <- bind_rows(bot_schools, bot_3)
}

```

Lets now visualize this

```{r}
ggplot(bot_schools, aes(x = as.factor(Year), y = Total_Crimes, fill = School)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Year", y = "Total Number of Crimes", title = "Bottom 3 Schools with Most Crime Each Year") +
  scale_fill_brewer(palette = "Set2") + 
  theme_minimal() +
  theme(legend.title = element_blank())  
```


# Average

## Which Schools Have the Most Crime Near them

```{r}

# to store the results
top_schools <- data.frame()


for (year in unique(final_df$Year)) {
  # Filter data for the current year
  year_data <- filter(final_df, Year == year)
  
  # Sum the crimes for each school
  year_sum <- year_data %>%
    summarise(across(3:38, mean)) %>%
    melt(id.vars = NULL, variable.name = "School", value.name = "Total_Crimes") %>%
    arrange(desc(Total_Crimes))
  
  # Select the top 3 schools
  top_3 <- year_sum[1:3, ]
  top_3$Year <- year
  
  top_schools <- bind_rows(top_schools, top_3)
}

```

Lets now visualize this

```{r}
ggplot(top_schools, aes(x = as.factor(Year), y = Total_Crimes, fill = School)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Year", y = "Average Number of Crimes Per Month", title = "Top 3 Schools with Most Crime Each Year") +
  scale_fill_brewer(palette = "Set3") + 
  theme_minimal() +
  theme(legend.title = element_blank())  
```
This is very useful information 


## Which Schools Have the Least Crime Near them

```{r}

# to store the results
bot_schools <- data.frame()


for (year in unique(final_df$Year)) {
  # Filter data for the current year
  
  year_data <- filter(final_df, final_df$Year == year)
  # Sum the crimes for each school
  year_sum <- year_data %>%
    summarise(across(3:38, mean)) %>%
    melt(id.vars = NULL, variable.name = "School", value.name = "Total_Crimes") %>%
    arrange(Total_Crimes)
  
  # Select the bottom 3 schools
  bot_3 <- year_sum[1:3, ]
  bot_3$Year <- year
  
  bot_schools <- bind_rows(bot_schools, bot_3)
}

```

Lets now visualize this

```{r}
ggplot(bot_schools, aes(x = as.factor(Year), y = Total_Crimes, fill = School)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Year", y = "Average Number of Crimes Per Month", title = "Bottom 3 Schools with Most Crime Each Year") +
  scale_fill_brewer(palette = "Set2") + 
  theme_minimal() +
  theme(legend.title = element_blank())  
```



